---
title: "Hands-on session 1"
author: "Montse Garcia Cerqueda (mgcerqueda@gmail.com)"
date: "Last update: `r format(Sys.time(), '%d %B, %Y')`"      
output:
  html_document:
    toc: true
    fig_caption: true
    number_sections: true
---

------------------------------------------------------------------------

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Analysis of the Heart Disease Dataset

Load the data from [here](https://raw.githubusercontent.com/jpinero/DMI_2021/main/datasets/heart_disease_dataset.csv), and the description is [here](https://raw.githubusercontent.com/jpinero/DMI_2021/main/datasets/heart_disease_description.txt). The original dataset comes from [here](https://archive.ics.uci.edu/ml/datasets/Heart+Disease) and corresponds to the [processed Cleveland data](https://archive.ics.uci.edu/ml/machine-learning-databases/heart-disease/processed.cleveland.data)

## Perform an EDA on the dataset

We will first start by reading the file containing our data.

```{r}
dataset <- read.csv(file = "data/heart_disease_dataset.csv", sep = " ", header = T, dec = ".", stringsAsFactors = F)
```

We can see our data has 303 rows, corresponding to individuals or patients, and 15 columns, one that identifies each individual and 14 different attributes. 
<details> 
<summary>The description of each attribute can be checked here:</summary>
-   **Age**: age in years.

-   **Sex**: sex (1 = male; 0 = female).

-   **Cp**: chest pain type.

    -   Value 1: typical angina.
    -   Value 2: atypical angina.
    -   Value 3: non-anginal pain.
    -   Value 4: asymptomatic.

-   **Trestbps**: resting blood pressure (in mm Hg on admission to the hospital).

-   **Chol**: serum cholestoral in mg/dl.

-   **Fbs**: fasting blood sugar \> 120 mg/dl (1 = true; 0 = false)

-   **Restecg**: resting electrocardiographic results.

    -   Value 0: normal.
    -   Value 1: having ST-T wave abnormality (T wave inversions and/or ST elevation or depression of \> 0.05 mV).
    -   Value 2: showing probable or definite left ventricular hypertrophy by Estes' criteria.

-   **Thalach**: maximum heart rate achieved.

-   **Exang**: exercise induced angina (1 = yes; 0 = no).

-   **Oldpeak**: T depression induced by exercise relative to rest.

-   **Slope**: the slope of the peak exercise ST segment:

    -   Value 1: upsloping.
    -   Value 2: flat.
    -   Value 3: downsloping.

-   **Ca**: number of major vessels (0-3) colored by flourosopy.

-   **Thal**: 3 = normal; 6 = fixed defect; 7 = reversable defect.

-   **Num**: diagnosis of heart disease (angiographic disease status).

    - Value 0: < 50% diameter narrowing.
    - Value 1: > 50% diameter narrowing.
</details>

### Data cleaning and missing data 
Once we have our data and we have familiarized ourselves with its attributes, we should check for any potential missing/incorrect data. To do so we can use the package `dplyr` and `fauxanif`, which includes functions such as `na_if_in()` or `na_if_not()`. We will "transform" any missing data, represented with a -9, as well as any data that we consider incorrect or improbable, with an `NA` value. 

```{r load-libraries, message=FALSE, warning=FALSE}
library(dplyr)
library(fauxnaif)

dataset <- dataset %>% mutate(
  age = na_if_not(age, 18:100), 
  across(c(sex, fbs, exang), na_if_not, c(0,1)),
  cp = na_if_not(cp, 1:4),
  trestbps = na_if_not(trestbps, 80:200),
  # ? chol = na_if_not()
  restecg = na_if_not(restecg, 0:2),
  thalach = na_if_not(thalach, 60:210),
  oldpeak = na_if_in(oldpeak, ~ . < 0),
  slope = na_if_not(slope, 1:3),
  ca = na_if_not(ca, 0:3), 
  thal = na_if_not(thal, c(3,6,7)),
  num = na_if_not(num, 0:4)
)

```

Afterwards, we will use the package `mice` to replace the missing values an estimated value. Before that, we can observe the missing data pattern using the function `md.pattern()`: 

```{r message=FALSE, warning=FALSE}
library(mice)
md.pattern(dataset,  rotate.names = TRUE)

```

## Create visualizations in order to show which variables seem to be more associated with heart disease

```{r}

```

# Difference in mortality rates in hospitalized COVID-19 patients

Using the supplementary material from the [Difference in mortality rates in hospitalized COVID-19 patients identified by cytokine profile clustering using a machine learning approach: An outcome prediction alternative](https://www.frontiersin.org/articles/10.3389/fmed.2022.987182/full), perform the following tasks

## Reproduce Figure 1 from the publication

```{r}

```

## Reproduce Figure 2 from the publication

but instead of representing the clusters in the annotation, represent the groups (G1 to G4)

```{r}

```

## Improve figure 2 of the publication

Add a second annotation with information of deathm and a third one with information of gender

```{r}

```

# Session info {.unnumbered}

```{r, results='asis',  echo=FALSE, message=FALSE }
sessionInfo()
```
